---
title: "Reproducible Research Assignment 1"
author: "Peter Cross"
date: "Sunday, February 28, 2016"
output: html_document
---

```{r}
library(dplyr)
library(lattice)
```

## Loading and preprocessing the data

Set the working directory to where the activity data is located
```{r, echo=TRUE}
setwd("c:/Users/peter/Downloads/repdata-data-activity/")
```



Read in the data
```{r, echo=TRUE}
activity <- read.csv("activity.csv")
```


Aggregate step data into daily levels, generate historgram and rug
```{r,echo=TRUE}
daily <- aggregate(steps ~ date, data=activity,FUN=sum, na.rm=TRUE)
hist(daily$steps,breaks=25)
rug(daily$steps)
```
## What is mean total number of steps taken per day?


Determine Median and Mean for daily step levels, removing NA values
```{r, echo=TRUE}
step.median <- median(daily$steps,na.rm=TRUE)
step.median
step.mean <- mean(daily$steps,na.rm=TRUE)
step.mean
```

## What is the average daily activity pattern?

Determine the Hourly means across the days, and plot the line graph
```{r,echo=TRUE}
hourly <- aggregate(steps ~ interval,data=activity,FUN=mean,na.rm=TRUE)
plot(hourly$steps ~ hourly$interval,type="l",xlab="Time of Day",ylab="Steps",main="Average Steps by Hour of Day")
```

determine the period of maximum mean steps
```{r,echo=TRUE}
max.hour <- subset(hourly,hourly$steps == max(hourly$steps))
max.hour$interval
```

## Imputing missing values

Determine the count of NA readings
```{r,echo=TRUE}
sum(is.na(activity$steps))
```

Determine the median steps for each interval
```{r,echo=TRUE}
median.interval <- aggregate(steps ~ interval,data=activity,FUN=median,na.rm=TRUE)
```

Append the median step level for each interval
```{r,echo=TRUE}
activity2 <- merge(activity,median.interval,by="interval",all.x=TRUE)
```

Determine positions of NA values in original dataset
```{r,echo=TRUE}
na.count <- is.na(activity2$steps.x)
```

replace any NA values in origianl data with median values for that interval
```{r,echo=TRUE}
activity2$steps.x[na.count] <- activity2$steps.y[na.count]
```

remove extraneous columns
```{r,echo=TRUE}
activity2 <- select(activity2,steps = steps.x,date,interval)
```

determine the new Histogram, daily values, and median and means for comparison
```{r,echo=TRUE}
daily2 <- aggregate(steps ~ date, data=activity2,FUN=sum, na.rm=TRUE)
hist(daily2$steps,breaks=25)
rug(daily2$steps)
step.mean2 <- mean(daily2$steps)
step.median2 <- median(daily2$steps)
step.mean2
step.median2
```

Compare the mean and median differences between datasets with NA values removed and NA values replaced.
```{r,echo=TRUE}
mean.diff <- step.mean - step.mean2
median.diff <- step.median - step.median2
mean.diff
median.diff
```

## Are there differences in activity patterns between weekdays and weekends?

Append a factor column for weekend/weekday
```{r,echo=TRUE}
activity2$date <- as.POSIXct(activity2$date,format='%Y-%m-%d')
activity2 <- mutate(activity2,"day_type" = as.factor(ifelse(weekdays(date) == "Saturday" | weekdays(date) == "Sunday","weekend","weekday")))
```

find means for each of the day types by interval and graph them
```{r,echo=TRUE}
#par(mfcol=c(2,1))
act2_mean <- aggregate(activity2$steps,by=list(activity2$interval,activity2$day_type),FUN=mean)
names(act2_mean) <- c("interval","day_type","steps")
xyplot(steps ~ interval | day_type,act2_mean,type="l",lwd=1,xlab="Interval",ylab="Number of steps",layout=c(1,2))
```

